A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-01-11 doc/large_strain.tex
extend Roofline discussion, add 2d figure and 3d (incomplete)

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.8\textwidth}
    \centering
    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline.pdf}
    % \caption{2D multigrid mesh with 2 global refinements}
  \end{subfigure}
  \caption{Roofline model for a selection of polynomial degrees on a 2D test example.}%
  \label{fig:roofline}
\end{figure}

{\color{red}
Figure \ref{fig:roofline} shows the Roofline performance model.
As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$ (4.5\% of the peak performance).
Matrix-free implementation of fine-strain elastic tangent operators (on average) leads to a much better performance of
$22.6\,\rm{GFlop/s}$, as well as about an order of magnitude higher computational intensity, but still stays in the memory bound regime.
The achieved performance is about 13\% of the peak arithmetic performance, which is considerably lower than the
70\% reported for the Laplace operation on Cartesian meshes in \cite{kronbichler12}.
When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),
we observe very marginal increase in computational intensity and almost unchanged performance.
This indicates that the origin for suboptimal results of matrix-free approach is most likely not related to the algorithms numerically investigated here.
Further detailed studies are necessary to explain this behavior.
}

{\color{red}
To summarize, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order bases to two factors.
First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
bound, see Figure \ref{fig:roofline}.
Second, is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
}

>>>>

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline_2d.pdf}
    \caption{2D}
    \label{fig:roofline_2d}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{LIKWID_Emmy_RRZE_roofline_3d.pdf}
    \caption{3D}
    \label{fig:roofline_3d}
  \end{subfigure}
  \caption{Roofline model for a selection of polynomial degrees. `Peak DP' denotes the peak double precision performance ($P$),
  `w/o FMA' represents a ceiling without Fused Multiply-Add ($P/2$), finally `w/o SIMD' is a ceiling where both FMA
  and SIMD vectorization are discarded ($P/8$).}%
  \label{fig:roofline}
\end{figure}

{\color{red}
Figure \ref{fig:roofline} shows results of the Roofline performance model.
As expected, the matrix-based approach is memory bound and (on average) achieves the performance of $8.5\,\rm{GFlop/s}$ (4.5\% of the peak performance) in 2D.
Matrix-free implementation of fine-strain elastic tangent operators (on average) leads to a much better performance of
$22.6\,\rm{GFlop/s}$ in 2D, as well as about an order of magnitude higher computational intensity, but still stays in the memory bound regime.
The achieved performance is about 13\% of the peak arithmetic performance. This is much lower than the
70\% reported for the Laplace operator on Cartesian meshes in \cite{kronbichler12}.
Unfortunately, the performance measurements for non-Cartesian meshes were not reported in \cite{kronbichler12}.

When comparing the data for ``tensor4'' caching strategy (Algorithm \ref{alg:mf_tensor4}) to that from the ``tensor2'' (Algorithm \ref{alg:mf_tensor2}),
we observe a marginal increase in computational intensity and almost unchanged performance.
This coincides with our expectation that the later algorithm requires transferring fewer bytes to perform operations on a single quadrature point.
On another hand, the ``scalar'' caching strategy (Algorithm \ref{alg:mf_scalar}) demonstrates a higher computational intensity.
Given that in this case we need to evaluate gradients with respect to both deformed and undeformed configurations, an additional
second order tensor per quadrature point (inverse of the Jacobian transformation) will be cached as compared to the Algorithm \ref{alg:mf_tensor2}.
As a result both algorithms will have comparable memory requirements (also confirmed by Figure ....),
however the ``scalar'' algorithm performs a large number of repetitive operations that are
independent on the right hand side vector (steps 8--11 in Algorithm \ref{alg:mf_scalar}), which makes it the worst option among the three.

Finally we note that all three algorithms are far from the theoretical peak performance which
indicates that the origin for suboptimal results of matrix-free approach is most likely not related to the various flavours of algorithms numerically investigated here.
Further detailed studies are necessary to explain this behavior.

To summarize this section, we can attribute the speedup of the matrix-free finite-strain tangent operators for higher order bases to two factors.
First, is the higher algorithmic intensity as compared to the matrix-based implementation, which is clearly memory
bound, see Figure \ref{fig:roofline}.
Second, is the reduction in the number of floating point operations per DoF thanks to the sum factorization technique, as discussed in
Section \ref{sec:mf} and \cite[Section 2.4]{kronbichler12}.
Among the three suggested implementations, the Algorithm \ref{alg:mf_tensor4} is the most flexible approach with relatively high computational intensity.
}
